---
title: "Data Visualization"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r,echo=FALSE,warning=FALSE, message=FALSE, comment=NA}
#libraries
library(readxl)
library(plotly)
library(tidyr) # Reshape the data for easier plotting
library(tidyverse)
library(ggplot2)
library(dplyr)
library(GGally) #scatter plot matrix
```




```{r,echo=FALSE, comment=NA, warning=FALSE, message=FALSE}
file_path <- "data/average_swims.xlsx"
metal_conc_data <- read_excel(file_path, sheet = 1)
```


```{r, echo=FALSE, comment=NA, warning=FALSE, message=FALSE}
summary(metal_conc_data[-1])

```

### Plots for variation of metal concentrations across each sample
<br>

```{r,echo=FALSE, fig.height=7, fig.width=12}

plot_ly(metal_conc_data, x = ~Sample_No) %>%
  # Add traces for each metal
  add_trace(y = ~Be, type = 'scatter', mode = 'lines+markers', name = 'Be') %>%
  add_trace(y = ~Cr, type = 'scatter', mode = 'lines+markers', name = 'Cr') %>%
  add_trace(y = ~Mn, type = 'scatter', mode = 'lines+markers', name = 'Mn') %>%
  add_trace(y = ~Fe, type = 'scatter', mode = 'lines+markers', name = 'Fe') %>%
  add_trace(y = ~Ni, type = 'scatter', mode = 'lines+markers', name = 'Ni') %>%
  add_trace(y = ~Cu, type = 'scatter', mode = 'lines+markers', name = 'Cu') %>%
  add_trace(y = ~As, type = 'scatter', mode = 'lines+markers', name = 'As') %>%
  add_trace(y = ~Se, type = 'scatter', mode = 'lines+markers', name = 'Se') %>%
  add_trace(y = ~Cd, type = 'scatter', mode = 'lines+markers', name = 'Cd') %>%
  add_trace(y = ~Sb, type = 'scatter', mode = 'lines+markers', name = 'Sb') %>%
  add_trace(y = ~Ba, type = 'scatter', mode = 'lines+markers', name = 'Ba') %>%
  add_trace(y = ~Tl, type = 'scatter', mode = 'lines+markers', name = 'Tl') %>%
  add_trace(y = ~Pb, type = 'scatter', mode = 'lines+markers', name = 'Pb') %>%
  add_trace(y = ~U, type = 'scatter', mode = 'lines+markers', name = 'U') %>%
  
  # Add dropdown to select the metal to display
  layout(
    title = 'Concentrations of Metals by Sample Number - (Dynamic plot)',
    xaxis = list(title = 'Sample Number'),
    yaxis = list(title = 'Concentration'),
    
    updatemenus = list(
      list(
        y = 0.8,
        buttons = list(
          # Add "All" button to display all metals
          list(method = "restyle",
               args = list("visible", rep(TRUE, 14)),
               label = "All"),
          list(method = "restyle",
               args = list("visible", c(TRUE, rep(FALSE, 13))),
               label = "Be"),
          list(method = "restyle",
               args = list("visible", c(FALSE, TRUE, rep(FALSE, 12))),
               label = "Cr"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, TRUE, rep(FALSE, 11))),
               label = "Mn"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, FALSE, TRUE, rep(FALSE, 10))),
               label = "Fe"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, FALSE, FALSE, TRUE, rep(FALSE, 9))),
               label = "Ni"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, rep(FALSE, 8))),
               label = "Cu"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, rep(FALSE, 7))),
               label = "As"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, rep(FALSE, 6))),
               label = "Se"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, rep(FALSE, 5))),
               label = "Cd"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, rep(FALSE, 4))),
               label = "Sb"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, rep(FALSE, 3))),
               label = "Ba"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE)),
               label = "Tl"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE, FALSE)),
               label = "Pb"),
          list(method = "restyle",
               args = list("visible", c(FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, FALSE, TRUE)),
               label = "U")
        )
      )
    )
  )

```


<br>

```{r, echo=FALSE, fig.height=8, fig.width=12, warning=FALSE, message=FALSE, comment=NA}


# Convert the data from wide to long format for plotting
metal_long <- metal_conc_data %>%
  pivot_longer(cols = Be:U, names_to = "Metal", values_to = "Concentration")

# Create a ggplot for the metals by Sample_No
p <- ggplot(metal_long, aes(x = Sample_No, y = Concentration)) +
  geom_line(aes(color = Metal), size = 1) +  # Line plot for concentration
  geom_point(aes(color = Metal), size = 1.5) +  # Add points on the lines
  facet_wrap(~ Metal, ncol = 4, nrow = 4, scales = "free_y") +  # Create grid with 4x4 layout
  theme_minimal() +  # Minimal theme for clean visualization
  labs(title = "Concentrations of Metals by Sample ",
       x = "Sample Number", 
       y = "Concentration") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),  # Center and size title
    strip.text = element_text(size = 10)  # Adjust facet label size
  )

# Print the plot
print(p)

```

<br>

### Box Plots of Metals - Identification of outliers and extreme points

<br>
<br>

```{r, echo=FALSE, fig.height=6, fig.width=10, warning=FALSE, message=FALSE, comment=NA}
# Convert the data from wide to long format for plotting
metal_long <- metal_conc_data %>%
  pivot_longer(cols = Be:U, names_to = "Metal", values_to = "Concentration")

# Create an interactive boxplot with a y-axis range slider
plot_ly(metal_long, 
        x = ~Metal, 
        y = ~Concentration, 
        type = 'box', 
        boxpoints = 'outliers',  # Show outliers
        jitter = 0.3,            # Add some jitter for visibility
        pointpos = -1.8,         # Control the position of the outliers
        text = ~paste("Sample No:", Sample_No)  # Show Sample No. when hovering
) %>%
  layout(
    title = 'Concentrations of Metals Across Samples - (Dynamic plot)',
    xaxis = list(title = ''),
    yaxis = list(title = 'Concentration'),


    # Add slider to control the y-axis range
    sliders = list(
      list(
        active = 0,
        currentvalue = list(prefix = "y-axis range: "),
        steps = list(
          list(label = "Full", method = "relayout", args = list("yaxis.autorange", TRUE)),
          list(label = "0-1000", method = "relayout", args = list("yaxis.range", list(0, 1000))),
          list(label = "0-500", method = "relayout", args = list("yaxis.range", list(0, 500))),
          list(label = "0-100", method = "relayout", args = list("yaxis.range", list(0, 100))),   
          list(label = "0-20", method = "relayout", args = list("yaxis.range", list(0, 20))),
          list(label = "0-5", method = "relayout", args = list("yaxis.range", list(0, 5))),
          list(label = "0-1", method = "relayout", args = list("yaxis.range", list(0, 1))),
          list(label = "0-0.2", method = "relayout", args = list("yaxis.range", list(0, 0.2)))
        )
      )
    )
  )

```


<br>
<br>

### Larval Activity Traces

<br>

```{r,echo=FALSE, comment=NA, warning=FALSE, message=FALSE}
larvel_activity_data <- read_excel(file_path, sheet = 2)
```

<br>



```{r, comment=NA,echo=FALSE,fig.width=8, fig.height=7,message=FALSE,warning=FALSE}
# Define function to filter data based on selected sample
filtered_data <- function(sample_num) {
  larvel_activity_data %>%
    filter(Sample == sample_num)
}

# Get unique sample numbers from the dataset
sample_numbers <- sort(unique(larvel_activity_data$Sample))

# Initialize plot with the first sample as default
plot <- plot_ly()

# Add traces for each fish in the first sample (default)
plot <- plot %>%
  add_lines(data = filtered_data(sample_numbers[1]),
            x = ~Time,
            y = ~AVG_Distance,
            color = ~factor(Fish),
            colors = "Set1",
            text = ~paste("Fish:", Fish, "<br>Time:", Time, "<br>Distance:", AVG_Distance),
            hoverinfo = 'text',
            name = ~paste("Fish", Fish))

# Add grey shaded areas for time intervals: 0-5, 10-15, 20-25
shaded_areas <- list(
  list(type = "rect",
       fillcolor = "lightgrey", opacity = 0.3, line = list(width = 0),
       x0 = 10, x1 = 15, xref = "x", y0 = 0, y1 = 1, yref = "paper"),
  list(type = "rect",
       fillcolor = "lightgrey", opacity = 0.3, line = list(width = 0),
       x0 = 20, x1 = 25, xref = "x", y0 = 0, y1 = 1, yref = "paper")
)

# Define the layout with dropdown
plot <- plot %>%
  layout(
    title = 'Larval Activity Over Time for each sample',
    xaxis = list(title = 'Time (minutes)'),
    yaxis = list(title = 'Average Distance Travelled'),
    shapes = shaded_areas,  # Add shaded areas
    updatemenus = list(
      list(
        type = 'dropdown',
        active = 0,
        buttons = lapply(sample_numbers, function(sample_num) {
          # For each sample, we dynamically update the plot with a new set of traces for each fish
          fish_data <- filtered_data(sample_num)
          unique_fish <- unique(fish_data$Fish)
          
          # Generate the new traces for each unique fish in the selected sample
          list(method = "restyle",  # Use 'restyle' to properly update the traces
               args = list(
                 # Update x and y data for each unique fish
                 list(
                   x = lapply(unique_fish, function(f) fish_data[fish_data$Fish == f,]$Time),
                   y = lapply(unique_fish, function(f) fish_data[fish_data$Fish == f,]$AVG_Distance),
                   text = lapply(unique_fish, function(f) paste("Fish:", f, "<br>Time:", fish_data[fish_data$Fish == f,]$Time, "<br>Distance:", fish_data[fish_data$Fish == f,]$AVG_Distance)),
                   # Update the name for each trace to update the legend
                   name = lapply(unique_fish, function(f) paste("Fish", f)),
                   hoverinfo = 'text'
                 )
               ),
               label = paste("Sample", sample_num))
        })
      )
    )
  )

# Display the plot
plot

```


<br>

* Substract __AVG_Distance__ by __AVG_Distance_Control__ to obtain the distances travelled by the larvae in the control sample, name this new variable as __Control_Distance__.


```{r, comment=NA,echo=FALSE,fig.width=8, fig.height=7,message=FALSE,warning=FALSE}
summary(larvel_activity_data[-c(1:5)])
```

<br>

Variables in the newly created data set:

* The mean distance traveled by the larvae in each time interval for every sample - __mean_D__
* The mean distance traveled by larvae in each time interval relative to the control - __mean_DC__
* The mean distance traveled in each time interval for the control sample - __mean_C__  


```{r, comment=NA,echo=FALSE,fig.width=8, fig.height=7,message=FALSE,warning=FALSE}

# Summarize the data by Sample and Time, and add Dark/Light column
summary_data <- larvel_activity_data %>%
  group_by(Sample, Time) %>%
  summarise(
    mean_D = mean(AVG_Distance, na.rm = TRUE),
    mean_DC = mean(AVG_Distance_Control, na.rm = TRUE),
    mean_C = mean(Control_Distance, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    `Dark/Light` = case_when(
      Time %in% c(5, 15, 25) ~ "D",
      Time %in% c(10, 20) ~ "L",
      TRUE ~ NA_character_  # Optional: Handle unexpected values
    )
  )

summary_data <- summary_data %>%
  select(1:2, last_col(), everything())

# Display the summarized data with Dark/Light column
head(summary_data,10)

```


```{r, comment=NA,echo=FALSE,fig.width=15, fig.height=9,message=FALSE,warning=FALSE}
p <- ggplot(summary_data, aes(x = Time, y = mean_D)) +
  geom_line(aes(color = Sample), size = 1) +  # Line plot for concentration
  geom_point(aes(color = Sample), size = 1.5) +  # Add points on the lines
  facet_wrap(~ Sample, ncol = 12, nrow = 10, scales = "free_y") +  # Create grid with 4x4 layout
  theme_minimal() +  # Minimal theme for clean visualization
  labs(title = "Mean Distance Travelled in each time interval by Sample Number - (Static plot)",
       x = "Sample Number", 
       y = "Mean Distance") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),  # Center and size title
    strip.text = element_text(size = 10)  # Adjust facet label size
  )

# Print the plot
print(p)

```


```{r, comment=NA,echo=FALSE,fig.width=15, fig.height=9,message=FALSE,warning=FALSE}
p <- ggplot(summary_data, aes(x = Time, y = mean_DC)) +
  geom_line(aes(color = Sample), size = 1) +  # Line plot for concentration
  geom_point(aes(color = Sample), size = 1.5) +  # Add points on the lines
  facet_wrap(~ Sample, ncol = 12, nrow = 10, scales = "free_y") +  # Create grid with 4x4 layout
  theme_minimal() +  # Minimal theme for clean visualization
  labs(title = "Mean Distance Travelled relative to the Control in each time interval by Sample Number - (Static plot)",
       x = "Sample Number", 
       y = "Normalized Mean Distance") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),  # Center and size title
    strip.text = element_text(size = 10)  # Adjust facet label size
  )

# Print the plot
print(p)

```

<br>
<br>


```{r, comment=NA,echo=FALSE,fig.width=8, fig.height=7,message=FALSE,warning=FALSE}

#Rename
colnames(larvel_activity_data)[colnames(larvel_activity_data) == "Sample"] <- "Sample_No"

# Summarize the larvel_activity_data to calculate ABO for each Sample_No
ABO_values <- larvel_activity_data %>%
  group_by(Sample_No) %>%
  summarise(
    ABO = sum(AVG_Distance_Control, na.rm = TRUE) / 24
  ) %>%
  ungroup()

# Join the calculated ABO values to metal_conc_data based on Sample_No
metal_conc_data <- metal_conc_data %>%
  left_join(ABO_values, by = "Sample_No")

```

### Correlation matrix of concentration data and aggregate behavioral distance  travelled

Created the "aggregate behavioral outcome" (ABO) variable in the metal_conc_data data set to produce the correlation matrix:

```{r, comment=NA,echo=FALSE,fig.width=15, fig.height=9,message=FALSE,warning=FALSE}

#summary_data <- summary_data %>% 
#                  group_by(Sample) %>%
#                  summarise(
#                    TMD = sum(mean_D, na.rm = TRUE),
#                    sum_mean_DC = sum(mean_DC, na.rm = TRUE),
#                    sum_mean_C = sum(mean_C, na.rm = TRUE)
#                  )

#summary_data <- summary_data %>%
#  rename(Sample_No = Sample)

#metal_conc_data <- metal_conc_data %>%
#  left_join(summary_data, by = "Sample_No")

```



```{r, comment=NA,echo=FALSE,fig.width=9, fig.height=7,message=FALSE,warning=FALSE}
ggpairs(metal_conc_data, columns = c(2:15, 17), aes(alpha = 0.55),   # Reduce alpha to range 0-1
        upper = list(continuous = wrap("cor", size = 2.5))) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1))
```



A low (0.2-0.35) positive linear correlation could be observed between ABO and:

*   U

A low (0.2-0.35) negative linear correlation could be observed between ABO and:

*   Ni, Cu, Cd, Ba and Pb

<br>

P-values for the correlation between metal and ABO.
```{r, comment=NA,echo=FALSE,fig.width=9, fig.height=7,message=FALSE,warning=FALSE}

selected_data <- metal_conc_data[, c(2:15, 17)]
p_values_table <- data.frame()

# Loop over each combination of variables and calculate the p-values
for(i in 1:(ncol(selected_data)-1)) {
  for(j in (i+1):ncol(selected_data)) {
    # Perform Pearson correlation test
    test <- cor.test(selected_data[[i]], selected_data[[j]], method = "pearson")
    
    # Store the variable names, correlation, and p-value
    p_values_table <- rbind(p_values_table, 
                            data.frame(Var1 = colnames(selected_data)[i], 
                                       Var2 = colnames(selected_data)[j], 
                                       p_value = test$p.value))
  }
}

# View the table of p-values
#print(p_values_table)

p_values_table %>% filter(Var2 == "ABO")

```

The metals Be, Ni, Cu, Cd, Ba, Tl, Pb and U have a P-value less than 0.15. 
